---
title: "SDS 220 Final Project - Written Report"
author: "Evelyn Dinh & Sophie Biderman"
format: pdf
editor: source
bibliography: citation.bib
csl: apa.csl
---

# Does Sleep Duration Influences Productivity and Stress Level of College Students?

```{r}
#| include: false
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(readr)

sleep_productivity_stress <- read_csv("sleep_productivity_stress.csv") |>
  clean_names() |>
  rename("early_bird_or_night_owl" = "are_you_more_of_an_early_bird_or_a_night_owl",
         "weekday_sleep_average" = "how_long_do_you_sleep_during_weekdays_on_average" ,
         "weekday_productivity_level" = "how_productive_are_you_during_weekdays_on_average_rate_your_productivity_on_the_following_scales_1_not_productive_2_4_somewhat_productive_5_very_productive",
         "weekday_stress_level" = "on_average_how_stressed_do_you_feel_during_weekdays_rate_your_stress_on_the_following_scales_1_not_stressed_2_4_somewhat_stressed_5_very_stressed")
  
sleep_productivity <- sleep_productivity_stress |>  
  select(weekday_sleep_average,
         weekday_productivity_level)

sleep_stress <- sleep_productivity_stress |>
  select(weekday_sleep_average,
         weekday_stress_level)
```

## Introduction

Imagine waking up after few hours of sleep, then dragging your tired body through hours of coursework, assignments, and even exams! Now imagine doing that mostly everyday, in a duration of months? According to @averagesleepduration and @recommendedsleep, most college students only get 6-6.9 hours of sleep per day on average while the recommended amount for college students is 7-9 hours. Even though it is understandable that there are certain deadlines, exams, or just work generally that seems to be more important than few extra hours of sleep, this constant trade-off may lead to unexpected consequences to both their physical and mental health. 

This project aims to investigate the relationship between sleep duration and 2 crucial aspects of any college student's life: their productivity and stress level. Our research question is **"Does the amount of sleep a college student gets influence their productivity and stress level?"**. As college students ourselves, our goal is to discover possible patterns of college students' sleep duration and whether it significantly accounts for our academic and work productivity and stress throughout the week. Understanding such connection, we can devise our own solutions to improve our college life inside-out, both academically and personally.

The data for this study was collected through a survey on Google Form. Our target population is college students, and our sample included students from different colleges. We mainly used a convenience sampling method, gathering many of our survey responses from our friends and acquaintances who are college students from Smith College, UMASS Amherst, Amherst College, Brown University, MIT...(USA), National Economic University, Diplomatic Academy of Vietnam, Hanoi University of Science and Technology... (Vietnam). 

## Exploratory Data Analysis

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep
#| fig-cap: "Weekday Sleep Duration"
#| echo: false

sleep_productivity_stress |>
  mutate(weekday_sleep_average = fct_relevel(weekday_sleep_average, 
            "Less than 5 hours", "5-7 hours", "More than 7 hours")) |>
  ggplot(aes(x = weekday_sleep_average,
             fill = weekday_sleep_average)) +
  geom_bar() +
  labs(
    title = "Weekday Sleep Duration of College Students",
    subtitle = "A Survey of Sleep Duration of College Students",
    caption = "Source: Self-Collected Survey",
    x = "Weekday Sleep Duration",
    y = "Count",
    fill = "Weekday Sleep Duration"
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: productivity
#| fig-cap: "Weekday Productivity Level"
#| echo: false

sleep_productivity_stress |>
  ggplot(aes(x = weekday_productivity_level,
             fill = weekday_productivity_level)) +
  geom_bar() +
  theme(
    axis.text = element_text(size = 7.5)
  ) +
  labs(
    title = "Weekday Productivity Level of College Students",
    subtitle = "A Survey of Productivity of College Students",
    caption = "Source: Self-Collected Survey",
    x = "Weekday Productivity Level",
    y = "Count",
    fill = "Weekday Productivity Level"
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: stress
#| fig-cap: "Weekday Stress Level"
#| echo: false

sleep_productivity_stress |>
  ggplot(aes(x = weekday_stress_level,
             fill = weekday_stress_level)) +
  geom_bar() +
  labs(
    title = "Weekday Stress Level of College Students",
    subtitle = "A Survey of Stress of College Students",
    caption = "Source: Self-Collected Survey",
    x = "Weekday Stress Level",
    y = "Count",
    fill = "Weekday Stress Level"
  )
```
## Results

### Sleep Duration vs. Productivity Level

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep_productivity
#| tbl-cap: "Sleep Duration vs. Productivity Level"
#| echo: false

sleep_productivity <- table(sleep_productivity)
knitr::kable(sleep_productivity)
```

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep_productivity
#| fig-cap: "Sleep Duration vs. Productivity Level"
#| echo: false

sleep_productivity_stress |>
  ggplot(aes(x = weekday_productivity_level,
             fill = weekday_sleep_average)) +
  geom_bar() +
  theme(
    axis.text = element_text(size = 7.5)
  ) +
  labs(
    title = "Sleep Duration vs. Productivity Level",
    subtitle = "A Survey of Sleep Duration vs. Productivity Level of College Students",
    caption = "Source: Self-Collected Survey",
    x = "Weekday Productivity Level",
    y = "Count",
    fill = "Weekday Sleep Duration"
  )
```

#### Hypotheses

- $H_0$: Sleep duration and productivity level are independent.
- $H_A$: Sleep duration and productivity level are related.

#### Conditions

1. **Independence**

2. **Large Sample**

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep_productivity
#| tbl-cap: "Sleep Duration vs. Productivity Level Expected Counts"
#| echo: false

chi_square_test <- chisq.test(sleep_productivity)
expected_counts <- chi_square_test$expected
knitr::kable(round(expected_counts, 2))
```

#### Chi-Square Testing

```{r}
#| echo: false
#| message: false
#| warning: false

print(chi_square_test)
```

### Sleep Duration vs. Stress Level

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep_stress
#| tbl-cap: "Sleep Duration vs. Stress Level"
#| echo: false

sleep_stress <- table(sleep_stress)
knitr::kable(sleep_stress)
```

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep_stress
#| fig-cap: "Sleep Duration vs. Stress Level"
#| echo: false

sleep_productivity_stress |>
  ggplot(aes(x = weekday_stress_level,
             fill = weekday_sleep_average)) +
  geom_bar() +
  labs(
    title = "Sleep Duration vs. Stress Level",
    subtitle = "A Survey of Sleep Duration vs. Stress Level of College Students",
    caption = "Source: Self-Collected Survey",
    x = "Weekday Stress Level",
    y = "Count",
    fill = "Weekday Sleep Duration"
  )
```

#### Hypotheses

- $H_0$: Sleep duration and stress level are independent.
- $H_A$: Sleep duration and stress level are related.

#### Conditions

1. **Independence**

2. **Large Sample**

```{r}
#| echo: false
#| message: false
#| warning: false

#| label: sleep_productivity
#| tbl-cap: "Sleep Duration vs. Stress Level Expected Counts"
#| echo: false

chi_square_test_2 <- chisq.test(sleep_stress)
expected_counts_2 <- chi_square_test_2$expected
knitr::kable(round(expected_counts_2, 2))
```

#### Chi-Square Testing

```{r}
#| echo: false
#| message: false
#| warning: false

print(chi_square_test_2)
```

## Discussion

## Appendix

## Bibliography
